GEE models

### 2017-07-01 doc/manuscript.Rmd
Moved figs/tabs to end, moved esm to own file

```{r figure2, dependson='geeData', fig.cap=fig_gee}
gee_results %>% 
    dplyr::filter(model == "Adjusted") %>% 
    plot_gee_main()
```

@@

@@

# Tables

```{r table1}
table_basic(ds, caption = tab_basic)
```

# Figure legends

```{r figure1, fig.cap=fig_dist}
plot_nefa_distribution(ds)
```

```{r figure2, dependson='geeData', fig.cap=fig_gee}
plot_gee_results(gee_results, leg = 'none')
```

# ESM Material

## ESM Methods

Generalized estimating equations (GEE) is analogous to mixed effects models, 
except it emphasizes population-level trends over individual-level trends that 
mixed models use. Because it can handle autocorrelations in the outcome (as 
would be seen in variables measured over time), GEE modeling is well suited to
longitudinal data. Latent class mixed models (LCMM) is a form of mixed effects
modeling that tries to categorize individuals into groups that have similar
clusters; in the case of the present analysis are the trajectories in beta-cell
function over time. Because mixed models use the participant as the random
effect, LCMM can identify individuals who similarly change over time.

Partial least squares (PLS) regression is a technique similar to principal
component analysis (PCA) that is designed to extract meaningful information from
multivariate, high dimensionality data (e.g. as in metabolomic and other -omic type 
analyses). These types of methods try to extract as much of the variance in the 
data into a smaller number of components or factors. The difference between PCA 
and PLS is that PLS is a supervised method, while PCA is not. This means that
PLS uses a response variable(s), i.e. an outcome or **Y**, to describe the
variation in the predictors (**X**) while PCA only describes the variation 
inherent in **X**. Because of this, PLS can be better at predicting the 
contributions of predictor variables against an outcome variable. Partial least
squares discriminant analysis (PLS-DA) is a form of PLS, except the **Y** is 
a discrete (i.e. categorical) variable and not a continuous one. Therefore, 
PLS-DA is able to predict who will belong to which group based on the values of
the predictors.

The two plots generated from the PLS-DA analysis show the results from both
the response (i.e. LCMM beta-cell function groups and dyslgycemia status) and
the predictors (all 22 NEFA). The response result plots as seen in 
`r cite_sf('plsda_groups')` and `r cite_sf('plsda_groups_dysgly')` show the
poor discriminatory ability of the PLS-DA model, with `r cite_sf('plsda_example')` 
showing a hypothetical example of good discriminatory ability. When the predictors represent an
underlying pattern that may exist between groups, the plots will show greater
separation between the groups in addition to a low misclassification rate. The
predictor result plots as seen in `r cite_sf('plsda_loadings')` and
`r cite_sf('plsda_loadings_dysgly')` show which of the predictor variables
contribute the most to the explained variance. Variables between the solid and 
dashed lines explain between 50-100% of the explained variance. Whether the 
variables have positive or negative values on either of the components axes 
determines how they contribute to each group. For instance, if those individuals
with dysglycemia have a positive value for component 1 and the fatty acid 16:0 also
has a positive value for component 1, this indicates that 16:0 is predictive of
having dysglycemia.

## ESM Tables and Figures

![`r sfig_consort`](../img/flowDiagramSample.png)

```{r ESM_Table1}
library(magrittr)
prep_qic <- gee_df %>%
    dplyr::select(
        SID, VN, lISSI2, lISI, TotalNE, Sex, Ethnicity, MET, ALT, BaseAge,
        Waist, BMI, AlcoholPerWk, TobaccoUse, FamHistDiab
        )

qic_is <- prep_qic %>% 
    dplyr::select(-lISSI2) %>% 
    stats::na.omit()

qic_bcf <- prep_qic %>% 
    dplyr::select(-lISI) %>% 
    stats::na.omit()

# QIC for IS
M0 <- geepack::geeglm(
    lISI ~ TotalNE + VN, data = qic_is,
    id = SID, family = stats::gaussian,
    corstr = 'ar1'
    )

M1 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex)
M2 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT) # best
M3 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET) # best
M4 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk) # best
M5 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab) # best
M6 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab + TobaccoUse) # best
M7 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TobaccoUse) # best
M8 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab) # best
M9 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TotalNE:VN)

qic_is_tab <- MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M9, rank = MuMIn::QIC) %>% 
    tibble::rownames_to_column() %>%
    dplyr::select(Model = rowname, QIC = IC, Delta = delta) %>%
    dplyr::mutate(id = seq_len(nrow(.))) %>%
    dplyr::bind_rows(., data.frame(Model = '**log(ISI)**', id = 0.5))

# QIC data for BCF
qic_bcf <- prep_qic %>% 
    dplyr::select(-lISI) %>% 
    stats::na.omit()

M0 <- geepack::geeglm(
    lISSI2 ~ TotalNE + VN, data = qic_bcf,
    id = SID, family = stats::gaussian,
    corstr = 'ar1'
    )
M1 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex)
M2 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT) 
M3 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET)
M4 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk)
M5 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab) # best
M6 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab + TobaccoUse) # best
M7 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TobaccoUse) # best
M8 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab) # best
M9 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TotalNE:VN) # best

qic_bcf_tab <- MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M9, rank = MuMIn::QIC) %>% 
    tibble::rownames_to_column() %>%
    dplyr::select(Model = rowname, QIC = IC, Delta = delta) %>% 
    dplyr::mutate(id = seq_len(nrow(.))+10) %>%
    dplyr::bind_rows(., data.frame(Model = '**log(ISSI-2)**', id = 10.5))

rbind(qic_is_tab, qic_bcf_tab) %>% 
    dplyr::arrange(id) %>%
    dplyr::select(-id) %>%
    dplyr::mutate(QIC = round(QIC, 1),
           Delta = round(Delta, 1)) %>% 
    pander::pander(missing = '', caption = stab_qic)
```

Given the number of possible combinations of outcome and predictor variables,
only ISI and ISSI-2 with total non-esterified fatty acids (nmol/mL) were used to
compare various GEE models and to select a final model.  Baseline age was used
as including both the original age and the time variable would result in
collinearity.  Column names are: QIC is the quasi-likelihood information criteria
(smaller values, eg. larger negative values, indicate a better fit compared to
other models), Delta is the QIC minus the lowest QIC (models with delta <10 are 
considered equivalent).  Models were:

- M0: log(ISSI-2) or log(ISI) = total non-esterified fatty acids (nmol/mL) + time
- M1: M0 + waist + sex + ethnicity + baseline age
- M2: M1 + ALT
- M3: M2 + physical activity
- M4: M3 + alcohol intake
- M5: M4 + family history of diabetes
- M6: M5 + smoking status
- M7: M2 + ALT + family history of diabetes + smoking status
- M8: M2 + ALT + family history of diabetes
- M9: M8 + time by NEFA interaction

```{r ESM_Table2}
table_distribution(ds, stab_dist)
```

```{r ESM_Fig2, fig.cap=sfig_corr}
plot_heatmap(ds)
```

```{r ESM_Fig3, fig.cap=sfig_gee}
plot_gee_results(gee_results_unadj)
```

```{r ESM_Table3_4}
table_gee(gee_results, stab_gee)
table_gee(gee_results_unadj, stab_gee_unadj)
```

```{r ESM_Fig4, fig.cap=sfig_lcmm}
plot_lcmm_results(lcmm_data)
```

```{r ESM_Fig5, fig.cap=sfig_plsda_groups}
plot_plsda_grouping(plsda_results, 'Beta-cell function')
```

```{r ESM_Fig6, fig.cap=sfig_plsda_loadings, fig.width=7, fig.height=7}
plot_plsda_loadings(plsda_results)
```

```{r ESM_Fig7, fig.cap=sfig_plsda_groups_dysgly}
plot_plsda_grouping(plsda_dysgly, 'Dysglycemia')
```

```{r ESM_Fig8, fig.cap=sfig_plsda_loadings_dysgly, fig.width=7, fig.height=7}
plot_plsda_loadings(plsda_dysgly)
```

```{r ESM_Fig9, fig.cap=sfig_plsda_example}
plot_plsda_grouping(example_plsda_results(), 'Example groups')
```

>>>>

@@

# Tables and figures

```{r table1_basic}
table_basic(project_data, caption = tab_basic)
```

```{r figure1_nefa, fig.cap=fig_dist}
plot_nefa_distribution()
```

```{r figure2_gee, dependson='geeData', fig.cap=fig_gee}
gee_results %>% 
    dplyr::filter(model == "Adjusted") %>% 
    plot_gee_main()
```

@@
